Causal Forest Approach for Site-specific Input Management via On-farm Precision Experimentation

### 2022-02-15 Writing/manuscript_cea.rmd

CF was developed specifically for identifying heterogeneous treatment effects [@athey2016recursive; @Wager2018a]. Like RF, CF works by building an ensemble of trees, called *causal trees* (CT), by recursively partitioning covariate space into two leaves, where each leaf estimates conditional average treatment effects as the mean differences between the dependent variable (here, yield) of the treated and control observations that belong to the leaf (see Appendix \@ref(ctree-ex) for an example of a causal tree that can help further clarify how CF works).

@@

The simulated field was created based on an actual 51-ha Illinois field on which the Data-Intensive Farm Management (DIFM) project [@bullock2019] has run multiple OFPEs. Figure \@ref(fig:field-map) illustrates that simulated trials consisted of $384$ 18.3 m $\times$ 73.2 m "plots," each of which was assigned an N fertilizer application rate. Each plot was made up of four (4-rows $\times$ 1-column) 18.3 m $\times$ 18.3 m "subplots," which were unit of analysis used in subsequent statistical analysis. Each subplot contained of a 6-rows $\times$ 6-columns grid of thirty-six 3.05 m $\times$ 3.05 m "cells." All field characteristics and yield data were generated at the cell level.

@@

Let $i \in \{1,2,\dots,1536\}$ index a simulated field's subplots and $j \in \{1,2,\dots,36\}$ index cells within subplots. Let $N_{i}$ be the N fertilization rate on subplot $i$. Eq. \@ref(eq:eqn6) shows the yield level on cell $j$ in subplot $i$:

\begin{equation}
f(N_{i}: \alpha_{i,j}, \beta_{i,j}, ymax_{i,j}) = ymax_{i,j}(1-exp(\alpha_{i,j} + \beta_{i,j} \cdot N_{i})) + \varepsilon_{i,j}. (\#eq:eqn6)
\end{equation}

The first term on the right-hand side of Eq. \@ref(eq:eqn6) is the deterministic yield component. The variable $N$ represents the nitrogen fertilizer application rate (kg/ha) which the trial designs varied by plot. Also in Eq. \@ref(eq:eqn6), $\alpha$, $\beta$, and $ymax$ are spatially varying parameters that govern how yield level is determined in conjunction with the applied N rate. They reflect local cell-level production conditions, such as soil characteristics and topography. For example, $\alpha$, which represents what the yield would be when no nitrogen is applied in conjunction with $ymax$, may reflect soil organic matter and residual nitrogen from previous year's production. $ymax$, which determines the maximum yield attainable, may include other micro nutrients like phosphorus and soil depth. $\beta$ can be interpreted as reflecting how efficiently the crop uses N to create grain mass, and might depend on soil properties that affect the availability of applied N fertilizer to the crop, say due to N immobilization, leaching, denitrification, and volatilization [@johnson2005nitrogen; @alva2006nitrogen]. The error term of the yield $\varepsilon$ represents the composite of all the unobserved factors that affect crop yield beyond $\alpha$, $\beta$, $ymax$ and $N$. Each cell $(i,j)$'s "true" optimal N application rate, denoted $N^{opt}_{i,j}$, was calculated by solving a profit maximization problem by cell assuming that the prices of corn and N fertilizer were $\$0.138/kg$ and $\$1.323/kg$, respectively. These prices are used consistently for our simualtions. See Appendix \@ref(par-gen) for further details on how the paremeters were generated and sample maps of the parameters and true economically optimal N rates.

@@

A cell-level dataset for the whole field, $\{(y_{i,j}, N^{opt}_{i,j}, N_{i,j}, \mathbf{c}_{i,j}): i \in \{1,2,\dots,1536\}, j \in \{1,2,\dots,36\}\}$ was generated by plugging in the parameter values, the error term, and the assigned nitrogen rates into \@ref(eq:eqn6). The cell-level data were then aggregated up to the subplot level by taking the mean of the cell-level values inside each subplot, which were the unit of observation. The subplot data were therefore

\begin{equation}
\{(y_{i}, N^{opt}_{i}, N_{i}) = \frac{1}{36}\sum_{j=1}_{36} (y_{i,j}, N^{opt}_{i,j}, N_{i,j}): i \in \{1,2,\dots,1536\}\}. (\#eq:eqn9)
\end{equation}

@@

In the third scenario, denoted "aabbyy," the researcher's dataset did not contain values of $\alpha_{i}$, $\beta_{i}$, or $ymax_{i}$. Rather, it contained subplot-level values of permutations of $\alpha$, $\beta$, and $ymax$. Each of the three parameters were split into two parts in a spatially correlated manner, and six new covariates were created (see Appendix \@ref(split-pars) for the details about this parameter-splitting process). These were $\alpha^{1}_{i}$, $\alpha^{2}_{i}$, $\beta^{1}_{i}$, $\beta^{2}_{i}$, $ymax^{1}_{i}$, and $ymax^{2}_{i}$. The two split variables were spatially correlated to reflect that many of the soil/field characteristics used to analyze OFPE data are spatially correlated. The values of $\alpha_{i}$, $\beta_{i}$ and $ymax_{i}$ did not indicate specific soil or field characteristics, but were functions of such. In reality, scientists cannot directly observe variables that can accurately predict yield plateau level, soil N content and N uptake efficiency. Rather, they use multiple observed soil/field characteristics to explain such phenological phenomena. For example, instead of using $ymax_{i}$ directly, some topographical land features and soil properties such as elevation and soil sand content could be used as yield-limiting factors [@jiang2004effect; @kravchenko2000correlation]. In reality, scientists may include more than three variables as covariates. This scenario should reduce the accuracy of EONR modeling for all the models compared to the ideal case.

@@

The RF, BRF, and CNN methods all followed the same conceptual steps to estimate site-specific EONR: 1) estimate the yield production functions and 2) calculate site-specific EONRs based on that estimated function. For the given modeling scenario, RF and BRF predict yield by using all the available covariates as explanatory variables in the estimation process. The `grf` package (version 1.2.0 [@tibshirani2018package]) in R was used for RF and BRF modeling. RF ann BRF from the `grf` package have several tunable hyperparameters (e.g., the minimum node size in each tree, the number of covariates used for node splitting, and parameters involving an honest tree-building process), and the optimal values for these parameters were selected by cross-validation. Individual trees in a forest were built with randomly subsampled observations, leading to a different prediction in every forest even when the same dataset was used. To reduce the variance of predictions for prediction accuracy, we set the number of trees (*num.trees*) at a high value, $2000$.

CNN leaves room for researchers to determine its architecture. We used a slightly modified version of one of the multi-stream CNN architectures proposed by @barbosa2020modeling, called "Late Fusion." Briefly stated, explanatory variables except N rate (e.g. $\alpha_{i,j}$, $\beta_{i,j}$, etc.) entered the model as a set of $6 \times 6$ element rasters.The input size of the N rate is treated as $1 \times 1$ because it is spatially homogeneous within each subplot. First, each input was connected to an independent convolutional layer with eight $3 \times 3$ filters each with stride one, followed by a $2 \times 2$ max-pooling layer with stride two. Then, a fully-connected rectified linear unit (ReLU) layer with sixteen neurons was added after each max-pooling layer, followed by a single ReLU neuron. Finally, multiple neurons were concatenated and fed to the fully-connected ReLU layer with sixteen neurons, followed by an output with a linear activation function. @barbosa2020modeling demonstrated that among several CNN architectures, this architecture modeled crop yield response to N rate management best. CNN was implemented in *Python* v3.7.6 using *Pytorch* v1.7.0 [@paszke2017automatic]. The Adam optimizer [@kingma2014adam] was used with a learning rate of $0.001 \%$ (default value). To avoid over-fitting, early stopping was used to monitor validation loss with a ten epochs of patience.

@@

\begin{equation}
\hat{N}^{opt}_{i} = \operatorname*{argmax}_{N}(p \cdot\hat{g}_m(N, \mathbf{\Omega_{i}}) - w \cdot N), (\#eq:eqn8)
\end{equation}

@@

\begin{equation}
\Delta Y_{N_1 \rightarrow N_m}(\mathbf{\Omega_{i}}) = \hat{\tau}_{N_{1} \rightarrow N_{m}}(\mathbf{\Omega_{i}}). (\#eq:eqn9)
\end{equation}

@@

Figure \@ref(fig:dist-optN) and Figure \@ref(fig:dist-piLoss) show the distributions of the one thousand RMSEs of the EONR estimations and $\hat{\pi}_{def}$ by ML methods and modeling scenarios, respectively (CNN is not shown as its RMSEs of the EONR estimations and $\hat{\pi}_{def}$ always took values $62.8$ kg/ha and \$$58.99$/ha, respectively.) The center of distribution mass of the CF-base's RMSE values and $\hat{\pi}_{def}$ are to the left of the centers of the distribution masses of the other methods' RMSE values, showing that CF-base more accurately and consistently estimated EONR across all the modeling scenarios. 
Interestingly, RF seems particularly vulnerable to the inclusion of more variables, with RMSE ($\hat{\pi}_{def}$) increasing from `r table_optN_prep[Model=="aby",rmse_optN_RF]` (`r table_optN_prep[Model=="aby", pi_loss_RF]`) in the "aby" scenario to `r table_optN_prep[Model=="aabbyytt", rmse_optN_RF]` (`r table_optN_prep[Model=="aabbyytt", pi_loss_RF]`) in the "aabbyytt" scenario. CF-base was more robust to the inclusion of additional variables. Overall, the simulations present CF-base as the clear winner in estimating EONR.

@@

Figure \@ref(fig:plot-tre) illustrates the underlying cause of CF-base's superior EONR predictions. It plots the estimated treatment effects against true treatment effects by treatment types and ML methods under the "aabbyytt" scenario in one of the one thousand simulations rounds. The figure shows points clustered around the red 1-to-1 line, meaning that the estimated treatment effects did not exhibit significant bias. However, the models differed substantially in the accuracy of their treatment effect estimations. Points are clustered more tightly about the 1-to-1 line for CF-base than RF and BRF. That is, CF-base estimated treatment effects much more efficiently than did RF and BRF, which focus on yield prediction rather than treatment effect estimation.

>>>>

CF was developed specifically for identifying heterogeneous treatment effects [@athey2016recursive; @Wager2018a]. Like RF, CF works by building an ensemble of trees, called *causal trees* (CT), by recursively partitioning covariate space into two leaves, where each leaf estimates conditional average treatment effects as the mean differences between the dependent variable (here, yield) of the treated and control observations that belong to the leaf (see Appendix A for an example of a causal tree that can help further clarify how CF works).

@@

The simulated field was created based on an actual 51-ha Illinois field on which the Data-Intensive Farm Management (DIFM) project [@bullock2019] has run multiple OFPEs. Figure \@ref(fig:field-map) illustrates that simulated trials consisted of 384 18.3 m $\times$ 73.2 m "plots," each of which was assigned an N fertilizer application rate. Each plot was made up of four (4-rows $\times$ 1-column) 18.3 m $\times$ 18.3 m "subplots," which were unit of analysis used in subsequent statistical analysis. Each subplot contained of a 6-rows $\times$ 6-columns grid of thirty-six 3.05 m $\times$ 3.05 m "cells." All field characteristics and yield data were generated at the cell level.

@@

Let $i \in \{1,2,\dots,1536\}$ index a simulated field's subplots and $j \in \{1,2,\dots,36\}$ index cells within subplots. Let $N_{i}$ be the N fertilization rate on subplot $i$. Eq. \@ref(eq:eqn3) shows the yield level on cell $j$ in subplot $i$:

\begin{equation}
f(N_{i}: \alpha_{i,j}, \beta_{i,j}, ymax_{i,j}) = ymax_{i,j}(1-exp(\alpha_{i,j} + \beta_{i,j} \cdot N_{i})) + \varepsilon_{i,j}. (\#eq:eqn3)
\end{equation}

The first term on the right-hand side of Eq. \@ref(eq:eqn3) is the deterministic yield component. The variable $N$ represents the nitrogen fertilizer application rate (kg/ha) which the trial designs varied by plot. Also in Eq. \@ref(eq:eqn3), $\alpha$, $\beta$, and $ymax$ are spatially varying parameters that govern how yield level is determined in conjunction with the applied N rate. They reflect local cell-level production conditions, such as soil characteristics and topography. For example, $\alpha$, which represents what the yield would be when no nitrogen is applied in conjunction with $ymax$, may reflect soil organic matter and residual nitrogen from previous year's production. $ymax$, which determines the maximum yield attainable, may include other micro nutrients like phosphorus and soil depth. $\beta$ can be interpreted as reflecting how efficiently the crop uses N to create grain mass, and might depend on soil properties that affect the availability of applied N fertilizer to the crop, say due to N immobilization, leaching, denitrification, and volatilization [@johnson2005nitrogen; @alva2006nitrogen]. The error term of the yield $\varepsilon$ represents the composite of all the unobserved factors that affect crop yield beyond $\alpha$, $\beta$, $ymax$ and $N$. Each cell $(i,j)$'s "true" optimal N application rate, denoted $N^{opt}_{i,j}$, was calculated by solving a profit maximization problem by cell assuming that the prices of corn and N fertilizer were $\$0.138/kg$ and $\$1.323/kg$, respectively. These prices are used consistently for our simualtions. See Appendix C.1 for further details on how the paremeters were generated and sample maps of the parameters and true economically optimal N rates.

@@

A cell-level dataset for the whole field, $\{(y_{i,j}, N^{opt}_{i,j}, N_{i,j}, \mathbf{c}_{i,j}): i \in \{1,2,\dots,1536\}, j \in \{1,2,\dots,36\}\}$ was generated by plugging in the parameter values, the error term, and the assigned nitrogen rates into Eq. \@ref(eq:eqn3). The cell-level data were then aggregated up to the subplot level by taking the mean of the cell-level values inside each subplot, which were the unit of observation. The subplot data were therefore

\begin{equation}
\{(y_{i}, N^{opt}_{i}, N_{i}) = \frac{1}{36}\sum_{j=1}^{36} (y_{i,j}, N^{opt}_{i,j}, N_{i,j}): i \in \{1,2,\dots,1536\}\}. (\#eq:eqn4)
\end{equation}

@@

In the third scenario, denoted "aabbyy," the researcher's dataset did not contain values of $\alpha_{i}$, $\beta_{i}$, or $ymax_{i}$. Rather, it contained subplot-level values of permutations of $\alpha$, $\beta$, and $ymax$. Each of the three parameters were split into two parts in a spatially correlated manner, and six new covariates were created (see Appendix C.2 for the details about this parameter-splitting process). These were $\alpha^{1}_{i}$, $\alpha^{2}_{i}$, $\beta^{1}_{i}$, $\beta^{2}_{i}$, $ymax^{1}_{i}$, and $ymax^{2}_{i}$. The two split variables were spatially correlated to reflect that many of the soil/field characteristics used to analyze OFPE data are spatially correlated. The values of $\alpha_{i}$, $\beta_{i}$ and $ymax_{i}$ did not indicate specific soil or field characteristics, but were functions of such. In reality, scientists cannot directly observe variables that can accurately predict yield plateau level, soil N content and N uptake efficiency. Rather, they use multiple observed soil/field characteristics to explain such phenological phenomena. For example, instead of using $ymax_{i}$ directly, some topographical land features and soil properties such as elevation and soil sand content could be used as yield-limiting factors [@jiang2004effect; @kravchenko2000correlation]. In reality, scientists may include more than three variables as covariates. This scenario should reduce the accuracy of EONR modeling for all the models compared to the ideal case.

@@

The RF, BRF, and CNN methods all followed the same conceptual steps to estimate site-specific EONR: 1) estimate the yield production functions and 2) calculate site-specific EONRs based on that estimated function. For the given modeling scenario, RF, BRF and CNN predict yield by using all the available covariates as explanatory variables in the estimation process. The `grf` package (version 1.2.0 [@tibshirani2018package]) in R was used for RF and BRF modeling. RF ann BRF from the `grf` package have several tunable hyperparameters (e.g., the minimum node size in each tree, the number of covariates used for node splitting, and parameters involving an honest tree-building process), and the optimal values for these parameters were selected by cross-validation. Individual trees in a forest were built with randomly subsampled observations, leading to a different prediction in every forest even when the same dataset was used. To reduce the variance of predictions for prediction accuracy, we set the number of trees (*num.trees*) at a high value, $2000$.

CNN leaves room for researchers to determine its architecture. We used a slightly modified version of one of the multi-stream CNN architectures proposed by @barbosa2020modeling, called "Late Fusion." Briefly stated, explanatory variables except N rate (e.g. $\alpha_{i,j}$, $\beta_{i,j}$, etc.) entered the model as a set of $6 \times 6$ element rasters.The input size of the N rate is treated as $1 \times 1$ because it is spatially homogeneous within each subplot. First, each input was connected to an independent convolutional layer with eight $3 \times 3$ filters each with stride one, followed by a $2 \times 2$ max-pooling layer with stride two. Then, a fully-connected rectified linear unit (ReLU) layer with sixteen neurons was added after each max-pooling layer, followed by a single ReLU neuron. Finally, multiple neurons were concatenated and fed to the fully-connected ReLU layer with sixteen neurons, followed by an output with a linear activation function. @barbosa2020modeling demonstrated that among several CNN architectures, this architecture modeled crop yield response to N rate management best. CNN was implemented in Python v3.7.6 using Pytorch v1.7.0 [@paszke2017automatic]. The Adam optimizer [@kingma2014adam] was used with a learning rate of $0.001 \%$ (default value). To avoid over-fitting, early stopping was used to monitor validation loss with a ten epochs of patience.

@@

\begin{equation}
\hat{N}^{opt}_{i} = \operatorname*{argmax}_{N}(p \cdot\hat{g}_m(N, \mathbf{\Omega_{i}}) - w \cdot N), (\#eq:eqn5)
\end{equation}

@@

\begin{equation}
\Delta Y_{N_1 \rightarrow N_m}(\mathbf{\Omega_{i}}) = \hat{\tau}_{N_{1} \rightarrow N_{m}}(\mathbf{\Omega_{i}}). (\#eq:eqn6)
\end{equation}

@@

Figure \@ref(fig:dist-optN) and Figure \@ref(fig:dist-piLoss) show the distributions of the one thousand RMSEs of the EONR estimations and $\hat{\pi}_{def}$ by ML methods and modeling scenarios, respectively. The center of distribution mass of the CF-base's RMSE values and $\hat{\pi}_{def}$ are to the left of the centers of the distribution masses of the other methods' RMSE values, showing that CF-base more accurately and consistently estimated EONR across all the modeling scenarios. 
Interestingly, RF seems particularly vulnerable to the inclusion of more variables, with RMSE ($\hat{\pi}_{def}$) increasing from `r table_optN_prep[type=="test" & Model=="aby",rmse_optN_RF]` (`r table_optN_prep[type=="test" & Model=="aby", pi_loss_RF]`) in the "aby" scenario to `r table_optN_prep[type=="test" & Model=="aabbyytt", rmse_optN_RF]` (`r table_optN_prep[type=="test" & Model=="aabbyytt", pi_loss_RF]`) in the "aabbyytt" scenario. CF-base was more robust to the inclusion of additional variables. Overall, the simulations present CF-base as the clear winner in estimating EONR.

@@

Figure \@ref(fig:plot-tre) illustrates the underlying cause of CF-base's superior EONR predictions. It plots the estimated treatment effects against true treatment effects by treatment types and ML methods under the "aabbyytt" scenario in one of the one thousand simulations rounds. Except in the cases of CNN, the figure shows points clustered around the red 1-to-1 line, meaning that the estimated treatment effects did not exhibit significant bias. However, the models differed substantially in the accuracy of their treatment effect estimations. Points are clustered more tightly about the 1-to-1 line for CF-base than RF, BRF and CNN. That is, CF-base estimated treatment effects much more efficiently than did RF and BRF, which focus on yield prediction rather than treatment effect estimation.
